Equalizer Programming Guide\\ {\footnotesize\sffamily\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-10-19 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\usepackage{tabularx,graphicx,a4,listings,wrapfig,subfigure}

@@

\title{Equalizer Programming Guide\\
  {\footnotesize\sffamily\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\\{\footnotesize Eyescale Software GmbH}
}
\date{
  \vspace{4cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
  \textbf{DRAFT}\\[\medskipamount]
  Version 0.9.1 for Equalizer v0.4\\[\medskipamount]
  \today
}

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9.1   & Oct 8, 2007  & review pass\\
    0.9     & Oct 7, 2007  & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Any questions related to Equalizer programming and this Programming
Guide should be directed to the \textsf{eq-dev} mailing
list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.

Equalizer is the next evolution of generic parallel programming
interfaces for visualization applications using OpenGL. Existing
solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
Cavelib\texttrademark\ and VRJuggler, implement a subset of concepts
similar to Equalizer. In other areas, e.g., tracking device support,
they provide more functionality.

@@

@@

\subsection{Compiling and running \textsf{eqPly}}

It is recommended to have a compiled version of Equalizer and the
examples when working with this Programming Guide. The Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
should be executed once to get familar with the capabilities of the
\textsf{eqPly} example.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
All functionality of Equalizer is accessed through the Equalizer client
library, which implements all functionality discussed in this
document. The client library provides communication between the Equalizer
processes.

\subsubsection{Server}

An Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster. Currently it
is only useful for running one application at a time, but it will be
extended to support multiple applications concurrently and efficiently
on one system. The server controls and potentially launches the
application's rendering clients.

@@

The application connects to an Equalizer server, which chooses a
configuration for the application. Furthermore, the application also
provides its render client, which will be controlled by the server. The
application reacts on events, updates its database and controls the
rendering.

@@

The render client implements --obviously-- the rendering part of an
application. Its execution is passive, that is, it has no main loop and
is completely driven by rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context.

The application might be a rendering client, in which case it can also
contribute to the rendering. If it does not implement any render
client-related code it is reduced to be the application's `master'
process without any OpenGL windows.

@@

The \textsf{eqHello} redraw function renders six rotating, colored quads
around the origin, where the viewer is usually positioned. The Equalizer
\textsf{frameDraw} meth\-od is used as a convience function to setup the
frustum and other OpenGL state. After setting up some additional
lighting parameter, \textsf{eqHello} rotates the scene and render the
quads using immediate mode:

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, that is, Equalizer imposes only the minimal, natural execution
framework upon the application. It does not provide a scene graph or
does interfere in any other way with the application's rendering
code. The restructuring work needed for Equalizer is needed anyway in
order to parallelize the application for rendering.

Methods called by the application have the form \textsf{verb[Noun]},
whereas methods called by Equalizer (`Task Methods') have the form
\textsf{nounVerb}. For example the application calls
\textsf{Config::startFrame} to render a new frame, which causes --among
other things-- \textsf{Node::frameStart} to be called in all active
render clients.

@@

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are in concept similar to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

\fig{fConfig} shows one example configuration for a four-side
CAVE\texttrademark, running on two machines (node) using three graphics
cards (pipe) with one window each to render to the four output channels
connected to the projectors for each of the walls. The compound
description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

For testing and development purposes it is possible to use multiple
instances for one resource, e.g., to run multiple render client nodes on
one computer. For deployment one node and pipe should be used for each
computer and graphics card, respectively.

@@

The root of the resource tree is the \textsf{eq::Config}, which
represents the current configuration of the application. The
configuration is the session in which all render clients are
registered. The config instance on a given node currently only holds the
local node, not all nodes of the configuration.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe, window
and channel task methods are executed from the pipe thread pipes or from
the main application thread for non-threaded pipes\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

An \textsf{eq::Window} holds a drawable and OpenGL context. The
drawable can be an on-screen window or an off-screen PBuffer or
FBO\footnote{off-screen drawables are not yet implemented, but can be
  created by the application and used with Equalizer}. The window holds
window-system-specific handles to the drawable and context, e.g., an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
its parent window. It is the entity executing the actual rendering. The
channel's viewport might be overwritten when it is rendering for another
channel.

@@

How the rendering resources are to be used is configured using a
compound tree. Although the API does not currently expose compounds, the
basic design behind compounds is explained here to increase the
understanding of the architecture behind Equalizer.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Frustrum descriptions}}
\end{wrapfigure}
The channels of the leaf compounds (source channels) in the compound
tree execute the rendering for the top-most compound's channel
(destination channel). The top-most compound with a channel defines the
2D pixel viewport rendered by all leaf compounds. The destination
channel and pixel viewport can not be overridden.

Compounds have a frustum description, which is inherited to the
children. The frustum can be overridden by children, but this is rarely
used since there is no use case for this behaviour.

The frustrum can be specified as a wall, which is completely defined by
the bottom-left, bottom-right and top-left coordinates relative to the
origin, or as a projection, which is defined by the position and
head-pitch-roll orientation of the projector, as well as the horizontal
and vertical field-of-view and distance of the projection
wall. \fig{fFrustra} illustrates the wall and projection frustum
parameters.

@@

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application of medium complexity. Its focus is not on rendering features
or visual quality, but it is used as a show case for the integration of
most of the Equalizer features.

@@

The node factory is used by Equalizer to create the object instances for
the rendering entities. Each of the classes inherits from the same type
provided by Equalizer in the \textsf{eq} namespace. The provided
\textsf{eq::NodeFactory} base class instantiates a 'plain' Equalizer
object, thus making it possible to selectively subclass individual
entity types, as it is done by \textsf{eqHello}. For each rendering
resource used in the configuration, one C++ object will be created:

@@

The third step is to create an instance of the application and to
initialize it locally. The application is an \textsf{eq::Client}, which
in turn is an \textsf{eqNet::Node}. The underlying network distribution
in Equalizer is a peer-to-peer network structure of
\textsf{eqNet::Node}s. The application programmer rarely is aware of the
classes in the \textsf{eqNet} namespace, but both the
\textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
initialization of nodes creates a local listening socket, so that the
node, and therefore the \textsf{eq::Client} can communicate over the
network with other nodes, such as the server and the rendering clients.

@@

After it has finished, the application and \textsf{eqPly} is
de-initialized and the \textsf{main} function returns:

@@

In the case of \textsf{eqPly} the application is also the render
client. The \textsf{eqPly} executable has three run-time behaviours:

@@

Finally it is time to initialize the configuration. For statistics, the
time for this operation is measures and printed. During initialization
the server launches and connects all render client nodes, and calls the
appropriate initialization task methods, as explained in later
sections. \textsf{Config::init} does return after all nodes, pipes,
windows and channels are initialized. It returns \textsf{true} only if
all initialization task methods were successful.

The \textsf{EQLOG} macro allows topic-specific logging. The numeric
topic values are specified in the respective \textsf{log.h} header
files, and logging for various topics is enables using the environment
variable \textsf{EQ\_LOG\_TOPICS}:

@@

When the frame is started, the server computes all rendering tasks and
sends them to the appropriate render client nodes. The render client
nodes dispatch the tasks to the correct node or pipe thread, were they
are executed in the order they arrive.

\begin{wrapfigure}{r}{.6\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames. This allows
overlapped execution in the node and pipe threads and minimizes idle
times. The first \textsf{nLatency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. \fig{fSyncAsync} shows the execution of (hypothetical) rendering
tasks without latency~(\fig{fSync}) and with a latency of
one~(\fig{fAsync}).

With \textsf{eqPly} a speedup of 15\% has been observed on a five-node
rendering
cluster when using a latency of one instead of no latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}.

When the main loop is finished, \textsf{Config::finishAllFrames} catches
up with the latency. It returns after all outstanding frames have been
rendered, and is needed here to provide an accurate measurement of the
framerate:

@@

The remainder of the application code cleans up in the reverse order of
the initialization. The config is exited, released and the connection to
the server is closed:

@@

In the second and third usage case of the \textsf{eqPly}, when the
executable is used as a render client, \textsf{Client::initLocal} never
returns. Therefore the application's main loop is never executed. In
order to keep the client resident, the \textsf{eqPly} example overrides
the client loop to keep it running beyond one configuration run:

@@

The \textsf{eqPly} example has a static distributed object to provide
initial data to all rendering nodes, as well as a versioned object to
provide frame-specific data, such as the camera position, to the
rendering methods.

@@

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData}, or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables can be outlayed
in one contiguous block of memory. The second approach is used
otherwise.

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and deserializing its data. The serialization is done in
\textsf{getInstanceData} and deserialization in
\textsf{applyInstanceData} by streaming all member variable to or from
the provided data streams. The data transport between nodes is
implemented in the data streams and various other Equalizer classes:

@@

The data output and input streams perform no type checking on the data
stream. It is the application's responsibility to match the order and
types of variables during serialization and deserialization exactly.

@@

Versioned objects have to override \textsf{isStatic} to return false, to
indicate that they are versioned. The current implementation has the
following characteristics:
\begin{itemize}
\item Only the master instance of the object is writable, that is,
  \textsf{eqNet::Object::com\-mit} can be called to generate a new
  version.
\item Slave instance versions can only be advanced, that is,
  \textsf{eqNet::Object::sync( version )} with a version smaller than
  the current version will fail.
\end{itemize}

@@

The frame data is registered first, since its identifier is transmitted
using the initialization data. The initialization data is registered
afterwards so that it can transmit the frame data's identifier. The
identifier of the initialization data is transmitted to the render
client nodes using the \textsf{initID} parameter of
\textsf{eq::Config::init}. Equalizer will pass this identifier to all
\textsf{configInit} calls of the respective objects:

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking device, as this is an orthogonal
problem to parallel rendering. Tracking device support has been solved
already by a number of implementations\footnote{VRCO Trackd, VRPN,
  etc.}, which can easily be integrated with Equalizer. The example code
in \textsf{eqPly} is one reference implementation for the integration of
such a tracking library:

@@

The exit of the configuration stops the render clients by calling
\textsf{eq::Config::exit}, and then deregisters the initialization and
frame data objects with the session:

@@

The rendering frames are issued by the application. The \textsf{Config}
only overrides \textsf{startFrame} in order to update its data before
forwarding the start frame request to the \textsf{eq::Config}.

@@

The camera position is updated and the frame data is commited, which
generates a new version for this object. This version is passed to the
rendering callbacks and will be used by the rendering threads to
synchronize the frame data to the state belonging to the current frame:

@@

Events are send by the render clients to the application using
\textsf{eq::Config::sendEvent}. At the end of the frame,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
the event handling. The default implementation processes all pending
events by calling \textsf{Config::handleEvent} for each of them.

@@

Foreach active render client, one \textsf{eq::Node} instance is
created on the appropriate machine. Nodes are only instantiated on their
render client processes, i.e., each process should have only one
instance of the \textsf{eq::Node} class. The application process might
also have a node class, which is handled in exactly the same way as the
render client nodes.

During node initialization the init data is mapped to a local instance
using the passed identifier from \textsf{Config::init}. The model is
loaded based on the filename in the initialization data. No pipe, window
or channel tasks methods are executed before \textsf{Node::configInit}
has returned.

@@

The node config exit deletes the loaded model and unmaps the
initialization data:

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entity
unlock all its children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

@@

The explicit synchronization of child or parent resources allows the
application to optimize the processing, by doing certain, independent
operations when the child or parent resources are already unlocked.

@@

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. The window and channel synchronization is similar and
omitted for simplicity. The \textsf{eqPly} example does not override
\textsf{Node::frameStart} or \textsf{frameFinish}, but it is absolutely
vital for the execution that \textsf{Node::startFrame} or
\textsf{Node::releaseFrame} are called, respectively. The default
implementation of the node task methods do take care of that.

@@

The config exit is again symmetric to the config initialization. The
frame data is unmapped and GPU-specific data is de-initialized by
\textsf{eq::Config::exit}:

>>>>

\usepackage{tabularx,graphicx,a4,listings,wrapfig,subfigure,textcomp}

@@

\title{Equalizer Programming Guide\\
  {\footnotesize\sffamily\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\\{\footnotesize Eyescale Software GmbH}
}
\date{
  \vspace{4cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
  \textbf{DRAFT}\\[\medskipamount]
  Version 0.9.2 for Equalizer v0.4\\[\medskipamount]
  \today
}

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9.2   & Oct 19, 2007 & Jonas' feedback, changes related to
                             per-node frame sync\\
    0.9.1   & Oct 8,  2007 & review pass\\
    0.9     & Oct 7,  2007 & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer is the next step in the evolution of generic parallel programming
interfaces for visualization applications using OpenGL. Existing
% rephrase ( aber ich weiss auch nicht wie :) )
solutions, such as SGI's OpenGL Multipipe\texttrademark\ SDK, VRCO's
Cavelib\texttrademark\ and VRJuggler, implement a subset of concepts
similar to Equalizer. In other areas, e.g., tracking device support,
they provide more functionality.

@@

If you have any question regarding Equalizer programming, this programming
guide, or specific problems you encountered, please direct them to the 
 \textsf{eq-dev} mailing list\footnote{see \link{http://www.equalizergraphics.com/lists.html}}.

@@

\subsection{Installing Equalizer and running \textsf{eqPly}}

Equalizer can be installed by downloading the
distribution\footnote{\link{http://www.equalizergraphics.com/downloads.html}}
and by compiling the source code. After installing Equalizer, please
take a look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of the \textsf{eqPly} example.

@@

The Equalizer architecture is based on a client-server model. Its client
library exposes all functionality discussed in this document to the
programmer, and provides communication between the different Equalizer
processes.

\subsubsection{Server}
Each Equalizer server is responsible for managing one visualization
system, i.e., a shared memory system or graphics cluster.  It controls
and launches the application's rendering clients.  Currently, Equalizer
currently only supports one application per server, but it will provide
concurrent and efficient multi-application support in future.

@@

The application connects to an Equalizer server and receives a
configuration.  Furthermore, the application also provides its render
client, which will be controlled by the server. The application reacts
on events, updates its database and controls the rendering.

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an
application. Its execution is passive, it has no main loop and
is completely driven by rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context.

The application might contain a rendering client, in which case it can 
also contribute to the rendering. If it does not implement any render
client-related code, it is reduced to be the application's `master'
process without any OpenGL windows.

@@

The \textsf{eqHello} redraw function renders six rotating, colored quads
around the origin. The \textsf{frameDraw} meth\-od provided by the
\textsf{eq::Channel} can be used as a convience function to setup the
frustum and other OpenGL state. After setting up some additional
lighting parameter, \textsf{eqHello} rotates the scene and render the
quads using immediate mode:

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, so Equalizer imposes only a minimal, natural execution
framework upon the application. It does not provide a scene graph, or
interfere in any other way with the application's rendering code. The
restructuring work required for Equalizer is needed anyway in order to
parallelize the application for rendering.

Methods called by the application have the form \textsf{verb[Noun]},
whereas methods called by Equalizer (`Task Methods') have the form
\textsf{nounVerb}. For example, the application calls
\textsf{Config::startFrame} to render a new frame, which causes --among
other things-- \textsf{Node::frameStart} to be called in all active
render clients.

@@

The application inherits from Equalizer classes and overrides virtual
functions to implement certain functionality, such as the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are similar in concept to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

\fig{fConfig} shows one example configuration for a four-side
CAVE\texttrademark, running on two machines (nodes) using three graphics
cards (pipes) with one window each to render to the four output channels
connected to the projectors for each of the walls. The compound
description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

For testing and development purposes it is possible to use multiple
instances for one resource, e.g. to run multiple render client nodes on
one computer. For deployment, one node and pipe should be used for each
computer and graphics card, respectively.

@@

%FIXME confusing
The root of the resource tree is the \textsf{eq::Config}, which
represents the current configuration of the application. The
configuration is the session in which all render clients are
registered. The config instance on a given node currently only holds the
local node, not all nodes of the configuration.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU).  In
the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. All pipe,
window and channel task methods are executed from the pipe's thread or
in the case of non-threaded pipes, from the main application
thread\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

An \textsf{eq::Window} holds a drawable and an OpenGL context. The
drawable can be an on-screen window or an off-screen\footnote{Off-screen drawables are not implemented yet, but can be
  created by the application and used with Equalizer} PBuffer or
FBO(Framebuffer object). The window holds
window-system-specific handles to the drawable and context, e.g. an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
its parent window. It is the entity executing the actual rendering. The
channel's viewport might be overwritten when it is rendering for another
channel during scalable rendering.

@@

Usage of the rendering resources is configured using a compound 
tree. Although the API does not currently expose compounds, the
basic design behind compounds is explained here for a better 
understanding of the Equalizer architecture.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Frustrum descriptions}}
\end{wrapfigure}
The channels of the leaf compounds in the compound tree are designated
as source channels, and the channel of the top-most compound is the
destination channel. Source channels execute the rendering for the
destination channel. The top-most compound with a channel defines the 2D
pixel viewport rendered by all leaf compounds. The destination channel
and pixel viewport can not be overridden.

Compounds have a frustum description, which is inherited to the
children. The frustrum can be specified as a wall, which is completely
defined by the bottom-left, bottom-right and top-left coordinates
relative to the origin, or as a projection, which is defined by the
position and head-pitch-roll orientation of the projector, as well as
the horizontal and vertical field-of-view and distance of the projection
wall. \fig{fFrustra} illustrates the wall and projection frustum
parameters.

@@

The \textsf{eqPly} example is shipped with the Equalizer distribution
and serves as a reference implementation of an Equalizer-based
application of medium complexity. Its focus lies not on rendering features
or visual quality, but on providing a working show case for the integration
of most of the Equalizer features.

@@

The node factory is used by Equalizer to create the objects or instances for
the rendering entities. Each of the classes inherits from the same type
provided by Equalizer in the \textsf{eq} namespace. The provided
\textsf{eq::NodeFactory} base class instantiates a 'plain' Equalizer
object, thus making it possible to selectively subclass individual
entity types, as it is done by \textsf{eqHello}. For each rendering
resource used in the configuration, one C++ object will be created:

@@

The third step is to create an instance of the application and to
initialize it locally. The application is an \textsf{eq::Client}, which
in turn is an \textsf{eqNet::Node}. The underlying network distribution
in Equalizer is a peer-to-peer network structure of
\textsf{eqNet::Node}s. The application programmer does not have to be
aware of the classes in the \textsf{eqNet} namespace, but both the
\textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
initialization of nodes creates a local listening socket, so that the
node, and therefore the \textsf{eq::Client} can communicate over the
network with other nodes, such as the server and the rendering clients.

@@

After finishing, the application and \textsf{eqPly} are
de-initialized and the \textsf{main} function returns:

@@

In the case of \textsf{eqPly}, the application is also the render
client. The \textsf{eqPly} executable has three run-time behaviours:

@@

Finally it is time to initialize the configuration. For statistics, the
time for this operation is measured and printed. During initialization
the server launches and connects all render client nodes, and calls the
appropriate initialization task methods, as explained in later
sections. \textsf{Config::init} does return after all nodes, pipes,
windows and channels are initialized. It returns \textsf{true} only if
all initialization task methods were successful.

The \textsf{EQLOG} macro allows topic-specific logging. The numeric
topic values are specified in the respective \textsf{log.h} header
files, and logging for various topics is enabled using the environment
variable \textsf{EQ\_LOG\_TOPICS}:

@@

When the frame is started, the server computes all rendering tasks and
sends them to the appropriate render client nodes. The render client
nodes dispatch the tasks to the correct node or pipe thread, where they
are executed in order of arrival.

\begin{wrapfigure}{r}{.6\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames. This allows
overlapping execution in the node and pipe threads and minimizes idle
times. The first \textsf{nLatency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. \fig{fSyncAsync} shows the execution of (hypothetical) rendering
tasks without latency~(\fig{fSync}) and with a latency of
one~(\fig{fAsync}).

With \textsf{eqPly}, a speedup of 15\% has been observed on a five-node
rendering
cluster when using a latency of one instead of no latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}.

When the main loop is finished, \textsf{Config::finishAllFrames} catches
up with the latency. It returns after all outstanding frames have been
rendered, and is needed to provide an accurate measurement of the
framerate:

@@

The remainder of the application code cleans up in the reverse order of
initialization. The config is exited, released and the connection to
the server is closed:

@@

In the second and third use case of the \textsf{eqPly}, when the
executable is used as a render client, \textsf{Client::initLocal} never
returns. Therefore the application's main loop is never executed. In
order to keep the client resident, the \textsf{eqPly} example overrides
the client loop to keep it running beyond one configuration run:

@@

The \textsf{eqPly} example has a static distributed object to provide
initial data to all rendering nodes, as well as a versioned object to
provide frame-specific data such as camera position to the
rendering methods.

@@

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData}, or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables can be stored
in one contiguous block of memory. The second approach is used
otherwise.

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and deserializing its data. Serialization in
\textsf{getInstanceData} and deserialization in
\textsf{applyInstanceData} is performed by streaming all member variable to or from
the provided data streams. The data transport between nodes is
implemented in the data streams and various other Equalizer classes:

@@

The data input and output streams perform no type checking on the data.
It is the application's responsibility to exactly match the order and
types of variables during serialization and deserialization.

@@

Versioned objects have to override \textsf{isStatic} to return false to
indicate that they are versioned. The current implementation has the
following characteristics:
\begin{itemize}
\item Only the master instance of the object is writable, that is,
  \textsf{eqNet::Object::com\-mit} can be called to generate a new
  version.
\item Slave instance versions can only be advanced, that is,
  \textsf{eqNet::Object::sync( version )} with a version smaller than
  the current version will fail.
\end{itemize}

@@

The frame data is registered before the initialization data, since its
identifier is transmitted using the \textsf{InitData}. The identifier of
the initialization data is transmitted to the render client nodes using
the \textsf{initID} parameter of \textsf{eq::Config::init}. Equalizer
will pass this identifier to all \textsf{configInit} calls of the
respective objects:

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking device, as this is an orthogonal
problem to parallel rendering. Tracking device support has been solved
already by a number of implementations\footnote{VRCO Trackd, VRPN,
  etc.}, which can easily be integrated with Equalizer. The example code
in \textsf{eqPly} provides a reference implementation for the integration of
such a tracking library:

@@

The exit function of the configuration stops the render clients by calling
\textsf{eq::Config::exit}, and then deregisters the initialization and
frame data objects with the session:

@@

The rendering frames are issued by the application. The
\textsf{eqPly::Config} only overrides \textsf{startFrame} in order to
update its data before forwarding the start frame request to the
\textsf{eq::Config}.

@@

The camera position is updated and the frame data is commited, which
generates a new version of this object. This version is passed to the
rendering callbacks and will be used by the rendering threads to
synchronize the frame data to the state belonging to the current frame:

@@

Events are sent by the render clients to the application using
\textsf{eq::Config::sendEvent}. At the end of the frame,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
the event handling. The default implementation processes all pending
events by calling \textsf{Config::handleEvent} for each of them.

@@

For each active render client, one \textsf{eq::Node} instance is
created on the appropriate machine. Nodes are only instantiated on their
render client processes, i.e., each process should only have one
instance of the \textsf{eq::Node} class. The application process might
also have a node class, which is handled in exactly the same way as the
render client nodes.

During node initialization the init data is mapped to a local instance
using the identifier passed from \textsf{Config::init}. The model is
loaded based on the filename in the initialization data. No pipe, window
or channel tasks methods are executed before \textsf{Node::configInit}
has returned.

@@

The node config exit function deletes the loaded model and unmaps the
initialization data:

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all its children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

@@

The explicit synchronization of child or parent resources allows the
application to optimize the processing, by performing certain, independent
operations when the child or parent resources are already unlocked.

The default thread synchronization synchronizes all
\textsf{Channel::frameDraw} operations on a \textbf{single} node with
the node's main thread. This facilitates porting, since the scene
database does not have to be multi-buffered as all threads of the node
process render the same frame.

The per-node frame synchronization does not break the asynchronous
execution across rendering nodes.  Advanced applications can remove per-node
frame synchronization.

The per-node frame synchronization is achieved through the
\textsf{startFrame\textrightarrow wait\-Fra\-me\-Started} and
\textsf{releaseFrameLocal\textrightarrow waitFrameLocal} synchronization
points. Note that this synchronization is only per-node, different nodes
in the cluster still run asynchronously.

The first synchronization ensures that the application (node) thread is
done modifiying the data, typically during event processing. The pipe
threads call \textsf{Node::waitFrameStarted} which blocks until the node
calls \textsf{startFrame} in \textsf{Node::frame\-Start}.

The second pair ensures that after \textsf{Config::finishFrame} all pipe
threads are done rendering the current frame, and therefore do not
access shared data. The node's \textsf{frameDrawFinish} waits for all
local pipes to release the synchronization by calling
\textsf{Pipe::releaseFrameLocal}, which happens by default in
\textsf{Pipe::frameDrawFinish}. The \textsf{frameDrawFinish} methods are
called after all \textsf{Channel::frameDraw} of the corresponding thread
have been executed.

@@

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. Please note that \textsf{Config::finishFrame} does
block until the current frame has been released locally and until the
frame \textsf{current - latency} has been released by all nodes. The
window and channel synchronization is similar and omitted for
simplicity.

It is absolutely vital for the execution that \textsf{Node::start\-Fra\-me}
and \textsf{Node::release\-Fra\-me} are called, respectively. The default
implementation of the node task methods does take care of that.

Since \textsf{eqPly} multi-buffers all dynamic data, it can completely
remove frame synchronization by:

\begin{itemize}
\item releasing the local synchronization in \textsf{Node::frameStart}
\item not calling \textsf{Node::waitFrameStarted} in
  \textsf{Pipe::frameStart}
\item not waiting for the pipe synchronization in \textsf{Node::frameDrawFinish}
\end{itemize}

Releasing the local synchronization early is done by calling
\textsf{releaseFrameLocal}, and \textsf{Node::frameDrawFinish} is
overwritten with an empty implementation:

{\footnotesize\begin{lstlisting}
void Node::frameDrawFinish( const uint32_t frameID,
                            const uint32_t frameNumber )
{ /* nop, see frameStart */ }

void Node::frameStart( const uint32_t frameID, const uint32_t frameNumber )
{
    startFrame( frameNumber ); // unlock pipe threads

    // Don't wait for pipes to release frame locally, sync not needed since all
    // dynamic data is multi-buffered
    releaseFrameLocal( frameNumber );
}
\end{lstlisting}}

@@

The config exit function is similar to the config initialization. The
frame data is unmapped and GPU-specific data is de-initialized by
\textsf{eq::Config::exit}:
